### 2025-01-09 skripsi/chapter4.tex
finish ch2 and start ch5

Using the subset of Fourier functions has several benefits which has motivated the choice. First these functions are characterized purely by their coefficients \(c_k\), meaning there is no need to store discretized values which potentially saves space and computation. The second benefit is that other functions such as polynomials can be approximated by them using the Fourier transform. This means that despite limiting the set of functions to Fourier functions, the behavior of PDEs with other sets of functions can be approximated to a certain extent. The final benefit is the mature ecosystem around these functions which include fast algorithms for the Fourier transform and even numerical approaches for solving known PDEs like the previously mentioned spectral method. In summary the generated data consists of features and labels which are Fourier functions implicitly defining the relationship enforced by a PDE and scenario. After establishing the kind of functions to be used in dataset generation, the process can proceed. There are 4 steps involved in the data generation:% TODO: add reference to second chapter for spectral method, fourier transform, and other external concepts.
\begin{enumerate}
  \item Scenario and PDE Determination:
        The first step of data generation is determining the scenario related to the PDE or governing equations. First, the PDE to be modeled is determined based on the goals of the dataset. Then, one or more of the parameters is chosen to predict the solution. The chosen parameters will be called the input functions and the solution will be called output functions from here on. The second part of the scenario is the domain or more simply the physical space occupied by the system to be modeled. The domain will be used to compute the function values in relation to the physical space from coefficients of Fourier functions.
  \item Parameter Determination:
        The second step is determining all parameters other than that input parameter based on the scenario and governing equation. These parameters may be coefficients such as material properties like density or viscosity, or forcing terms which model external influence on the system like a heat source in the case of the heat equation. Depending on the parameters, solutions of PDEs may behave very differently, such as the appearance of discontinuities in the solution to low viscosity Burgers' equation. Because of this, the choice of parameters is guided by what the dataset seeks to do.
  \item Random Coefficient Generation:
        The third step of data generation is randomly generating the solutions for the chosen equation. Generating random functions in the space of Fourier functions exploits the fact that the coefficients characterize the function completely. By randomly assigning coefficients \(c_k\), many functions can be generated randomly with very little cost. Since the coefficients need to be complex numbers as in \lccref{eq:complex_number}, both real and imaginary components are generated independently by assigning a random value to each component for each wave number \(k\). They are then put together again into complex numbers.

@@

\subsection{Data Generation: Anti-derivative}
\noindent The first dataset is a simple one dimensional derivative. This was chosen as a simple proof of concept of the ability to solve a differential equation. The equation is related to many real-world problems such as acceleration and speed. One can imagine a train in an ideal world where acceleration is directly translated into speed. When the train accelerates at time \(t \) by some amount \(a \), we can expect the train to have some speed \(u \). In this ideal world, the relationship between speed and acceleration can be modeled with a simple differential \lccref{eq:derivative}. This scenario is found in many real systems albeit often with many more details such as different components of acceleration from friction, drag, gravity, and other factors.
As previously mentioned in \lccref{sec:data_generation}, both velocity \(u \) and acceleration \(a \) are modeled with Fourier series in \lccrefs{eq:fourier_speed,eq:fourier_acceleration} respectively.
\begin{align}
  \dv{u\left( t \right)}{t} & = a\left( t \right) \label{eq:derivative}                         \\
  u\left( t \right)         & = \sum_{k} \hat{u}_k e^{2\pi ikt} \label{eq:fourier_speed}        \\
  a\left( t \right)         & = \sum_{k} \hat{a}_k e^{2\pi ikt} \label{eq:fourier_acceleration}
\end{align}
The domain of the scenario is a two-hour time window. This number was chosen because it is around the ideal length of travel time on high speed rail in comparison to air travel and car travel \autocite{givoniDevelopmentImpactModern2006,wangEfficiencySpatialEquity2019,wrro2236}. This is the first step in generating this dataset.

In the second step, as the derivative \lccref{eq:derivative} does not contain any parameters other than the acceleration which is the input parameter, there are no other parameters to determine. Therefore, the data generation process proceeds to generating coefficients for the speed functions \(\hat{u} \). The coefficients are assigned randomly from a Gaussian distribution with a mean of zero and standard deviation of one. This choice was made such that most wave numbers will have a coefficient of close to zero leaving a sparse set of wave numbers to mostly affect the resulting function. In total, 5000 unique functions are generated with 100 complex coefficients each.

@@

In the final step, using randomly generated values of \(\hat{a}_k\), the corresponding values of \(\hat{u}_k\) are computed with \lccref{eq:derivative_coeff}. These coefficients are then used to compute the function values inside the domain. The two-hour time window is represented by a grid of 500 discrete points. With the coefficients and evaluation points ready, the function values are computed using \lccrefs{eq:fourier_speed,eq:fourier_acceleration}. Next, noise is added to the function values in order to motivate models learning on the dataset to generalize on noise. To also allow evaluation of how well the model performs with different levels of noise, the samples are duplicated into three copies for each high, medium and low noise levels. The function values of each copy is perturbed with noise from a Gaussian distribution with zero mean and standard deviation of some percentage of the average function value standard deviation. The percentages of each high, medium, and low noise levels are 5, 10, and 50 percent. The perturbed function values are then transformed back into their coefficients. An example generated function and its different perturbed versions is shown in \lccref{fig:antiderivative_noise_levels}. The horizontal axis indicates time which is displayed in units of hours. The vertical axis indicates velocity in abstract units.

\begin{figure}[H]
  \centering
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_clean_perturbed_solution.pgf}
    \end{adjustbox}
    \caption{}\label{fig:antiderivative_noise_levels}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_clean.pgf}
    \end{adjustbox}
    \caption{}\label{fig:antiderivative}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_derivative_clean.pgf}
    \end{adjustbox}
    \caption{}\label{fig:derivative}
  \end{subfigure}
  \caption{(\subref{fig:antiderivative_noise_levels}) A generated function with no noise (clean), low noise level (5\%), medium noise level (10\%), and high noise levels (50\%).\ (\subref{fig:antiderivative}) Antiderivative function with no noise.\ (\subref{fig:derivative}) Derivative function with no noise.}
\end{figure}
% TODO: interesting note that model trains better on function to coefficient task when the antiderivative is computed from the randomly generated derivative than if the derivative is computed from the randomly generated antiderivative. THIS NEEDS A REASON AND ANALYSIS SO MIGHT NOT BE NEEDED IN THIS WRITING. MAYBE A GOOD IDEA FOR A SEPARATE PIECE

\subsection{Data Generation: Burgers' Equation}
\noindent The second dataset generated in this study concerns the Burgers' equation. This equation has been used to model a variety of cases including fluid dynamics, traffic flow, and, shock waves \autocite{bonkileSystematicLiteratureReview2018,orlandiBurgersEquation2000,becBurgersTurbulence2007,jamesonEnergyEstimatesNonlinear2007}. This equation is also used as a base problem for testing the effectivity of numerical methods in solving non-linear PDEs \autocite{barterShockCapturingPDEbased2008,banksNumericalErrorEstimation2012, tabatabaeiImplicitMethodsNumerical2007,bonkileSystematicLiteratureReview2018}. The nonlinear term in the equation creates steep gradients and even shock waves which are discontinuous with low viscosity conditions. These challenges test the stability of numerical solvers. This is the scenario and reason for the choice of this PDE\@. To control the viscosity and therefore the steepness of gradients, the formulation of the Burgers' equation considered in this study is the forced viscous Burgers' equation in one dimension. For a velocity of \(u(x, t)\), viscosity of \(\nu \), and forcing term of \(f(x, t)\), the formulation of the forced Burgers' equation is shown in \lccref{eq:forced_viscous_burgers}.
\begin{align}
  \pdv{u}{t}+u\pdv{u}{x}-\nu\pdv[2]{u}{x} & = f \label{eq:forced_viscous_burgers}
\end{align}
The domain of we consider here is based on an exact solution of the equation. This is done so that the learned solver model can be tested on the exact solution. There are a number of sources in literature on the exact solution of the Burgers' equation \autocite{bentonTableSolutionsOnedimensional1972,wazwazPartialDifferentialEquations2010,woodExactSolutionBurgers2006}. In this study, we use a specific solution by \textcite{woodExactSolutionBurgers2006} which can be seen in \lccref{eq:burgers_exact_solution}. This solution is periodic in space but not in time. Therefore, the domain in space will be based on the spatial periodicity of the exact solution which is two. As such, the space domain spans from 0 to 2. The time domain, on the other hand, is chosen to span from 0 to 10 such that enough details of the solution is included.
\begin{equation}
  u(x,t) = \frac{2\nu\pi e^{-\pi^2\nu t}\sin(\pi x)}{a+e^{-\pi^2\nu t}\cos(\pi x)} \label{eq:burgers_exact_solution}
\end{equation}
Continuing to the second step, the parameters involved in \lccref{eq:forced_viscous_burgers} are the viscosity \(\nu \) and the forcing term \(f \). Since the forcing term is dependent on the solution in this step, the only parameter to determine is the viscosity. We choose three constant viscosity values which are 0.1, 0.01, and 0.0. These values are chosen so that a variety of behaviors are represented in the dataset from viscous to inviscid flow.

After this, in the third step, the solution functions \(u \) are generated with 8 modes in time and 8 in space and 500 unique function samples for each viscosity value. Since we don't factor the viscosity in at this point, for ease of comparison, we ensure the same solutions are generated every time by using a new instance of the generator with the same seed number each time. This results in the solutions being the same for all viscosity and other parameters of the equation are adapted to this.

@@

For the reanalysis dataset, we limit the time range to the years 1995{-}2013 for training and 2020{-}2022 for testing. The specific time ranges for training are chosen to balance the diversity within the data and capacity of the machine the processing will be done on. The testing set time range was chosen to be as close to those shown on the WeatherBench 2 website. The filter for time and variables is shown in \lccref{fig:wb2_retrieval}. The reason for the different time ranges between training and testing is to reduce the risk of information leakage accross the training and testing sets. This ensures that the testing scores we measure will be for the most part from the generalization that model learned during training. The filter is an array the length of the number of samples. The array values are all the datetime values that the year ranges are true for. This results in an array of boolean values. Using this boolean filter we can index the data so that we get the filtered data. Then, the filtered data is saved locally so that we can access it in a more performant way. The encoding chunks may need to be removed in order to evade the Xarray library from mistakenly persisting the remote encoding after slicing data \autocite{ianZarrEncodingAttributes}.
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
  years_idx = ds['time.year']
  data_idx = (years_idx > START_YEAR) & (years_idx < END_YEAR)
  da_orig = ds['2m_temperature'][data_idx]
  # save locally
  da_orig.encoding.pop('chunks',None)
  da_orig.to_zarr(DATA_FILENAME, mode='w')
  # open local version
  da = xr.open_zarr(DATA_FILENAME)
  \end{lstlisting}
  \caption{Example of filtering of Xarray dataset by year and variable. The data is saved and then loaded locally.}\label{fig:wb2_retrieval}
\end{figure}

The dataset includes 62 variables with some variables also spanning 13 discrete vertical levels in the atmosphere. For our use, we will only be using the 2-meter temperature of the atmosphere. This is the air temperature 2 meters above the surface be it land, sea, or inland water. This variable is in units of kelvin (K). Vertically, the values of this variable is obtained by interpolating between the model values at the surface and the lowest model vertical level. An example of the data that was retrieved can be seen in \lccref{fig:wb2_data}. The data shown is in the equiangular projection. This means that towards the poles, the shapes become more distorted. While this is not the most ideal way to represent the data, using the data in this way adds to the challenge of how the model can adapt to the use of suboptimal configurations in the data. To compute the anomaly, we can simply subtract the current state in \lccref{fig:wb2_train_data} by the climatology in \lccref{fig:wb2_clim_data} resulting in \lccref{fig:wb2_train_anomaly}.
\begin{figure}[H]
  \centering
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_train_data.pgf}
    \end{adjustbox}
    \caption{}\label{fig:wb2_train_data}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_clim_data.pgf}
    \end{adjustbox}
    \caption{}\label{fig:wb2_clim_data}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_train_anomaly.pgf}
    \end{adjustbox}
    \caption{}\label{fig:wb2_train_anomaly}
  \end{subfigure}
  \caption{Data retrieved:(\subref{fig:wb2_train_data}) 2-meter temperature at January 1, 1995 6:00 UTC\@. (\subref{fig:wb2_train_anomaly}) Anomaly for 2-meter temperature at January 1, 1995 6:00 UTC\@. (\subref{fig:wb2_clim_data}) 2-meter temperature climatology for the first day of the year at hour 6 UTC\@.}\label{fig:wb2_data}
\end{figure}

After filtering, our training set has the dimensions of 27760 time steps, 64 longitude, and 32 latitude. The testing set on the other hand has 4384 time steps and the same values for the other two dimensions. For uses, such as a validation set during hyperparameter optimization, one can use a subset of the training set. The dimensions of the climatology are 4 for hour (0, 6, 12,\& 18), 366 for day of year, 64 for longitude, and 32 for latitude.

@@

The second use of a dataset depends on whether it can be interpreted as a mapping of a previous function to the next function. In other words, a series of functions related to the previous function as a sum of the previous function and a perturbation. A commonly used form of this are time series. Generally, a time series can be represented as \lccref{eq:time_step_function}. This function describes the evolution of function \(f\) as a sum of some initial state \(f^0\) and perturbation which is its rate of change \(\dv{f}{t}\) at a point in time close to the current time multiplied by the time step \(\delta t\). This is the first order Taylor polynomial. Adding more terms with higher order derivatives can be done to increase the accuracy. This describes the same process that is used in the finite difference method.
\begin{equation}
  f^{t+1} = f^{t} + \delta t \times \dv{f}{t} + \dots \label{eq:time_step_function}
\end{equation}
In this case the previous function and any other independent variables like the forcing term involved in the right-hand side of the equation is used to predict the difference between the future state and the current state. It would mean that each sample function in the data would need to at least have two dimensions. This is because one dimension represents the evolution of the function over time, while the other dimensions define the spatial domain where the function's values are computed at each time step. This condition is true for both the Burgers' equation dataset and the WeatherBench 2 collection of ERA5 datasets. First, the Burgers' equation dataset, while is also able to be used in the first way as a mapping between the forcing term and the solution, one of the traditional ways of solving it is using the method of lines. This entails solving the dimensions other that time with methods such as the finite differences method or the spectral method. The time dimension is then separately solved using the finite differences method such as the form shown in \lccref{eq:time_step_function} \autocite{schiesserNumericalMethodLines2012,sadikuSimpleIntroductionMethod2000}. This is the form that the Burgers' equation dataset will be used in. In this case, algebraic manipulation of \lccref{eq:forced_viscous_burgers} results in the time derivative of the solution as in \lccref{eq:forced_viscous_burgers_time_derivative}.
\begin{align}
  \pdv{u}{t} = f - u\pdv{u}{x} + \nu\pdv[2]{u}{x} \label{eq:forced_viscous_burgers_time_derivative}
\end{align}
The most important component of the right-hand side of the equation, the time derivative, is expressed in terms of the solution function, the forcing term and any other partial derivatives. Because of this, it means we can train the model to learn the relationship between the solution function combined with the forcing term and the time derivative. Putting this together with \lccref{eq:time_step_function}, it is possible to predict the solution function in a future time based on the current solution function and forcing term. The input functions in this case are both the value of the solution function and the forcing term at the current time step. The output function is the time derivative which is computed by rearranging \lccref{eq:time_step_function} resulting in \lccref{eq:time_derivative_time_step_function}.
\begin{equation}
  \dv{f}{t} = \frac{f^{t+1} - f^{t}}{\delta t} \label{eq:time_derivative_time_step_function}
\end{equation}
Perhaps in a more obvious manner, the same is also applied to the WeatherBench 2 collection of ERA5 datasets. The 2-meter temperature is a series of snapshots in time of the distribution of temperature all over the world at 2 meters above the surface. The input function in this case is the temperature distribution at the current time step. The output function is the time derivative of the temperature distribution computed using \lccref{eq:time_derivative_time_step_function}.

The input and output functions are then transformed into basis coefficients using the Fourier transform. % TODO: add reference to fourier transform in chapter 2
Once the coefficients are computed, the input and output sample pairs are split into train, test, and validation subsets. For the first type of input and output function pair, since the samples are independent of each other, a random sampling of the dataset without replacement is done to select the samples for each subset. For the second type of input and output function pairs, if there are multiple samples of functions as in the Burgers' equation dataset, the same process with random sampling is followed. On the other hand, if only one function sample is available, such as the WeatherBench 2 dataset, an issue that arises is the fact that the samples are dependent on one another. This is because a previous time step influences the next time step. This is a problem for generalizing the learned model, since random sampling would lead to the model learning information in the future of the testing subset \autocite{kapoorLeakageReproducibilityCrisis2023,kaufmanLeakageDataMining2012}. As such, the portioning of the dataset is done by dividing the sample into different sections of the function in time.

@@

\section{Implementation of Computational Model}
\noindent This section discusses the implementation of modeling and approximation of PDEs using LSSVR\@. Using the computational model design in \lccref{sec:computational_model_design}, the computational model is implemented as a library using the Python programming language. The development process itself is based on the iterative development model as mentioned in \lccref{sec:development method}. In this section, the output iterative processes are consolidated and explained together.

@@

\section{Experimental Results}\label{sec:experimental_results}
\noindent In this section, the results of each experimental scenario is presented and analyzed. The results come in the form of testing metrics mentioned in . %TODO: add reference to the metrics used.
The model interpretations are displays of the correlation image and the p-matrix. The function values and coefficients themselves will also be plotted depending on the number of dimensions.

>>>>

Using the subset of Fourier functions has several benefits which has motivated the choice. First these functions are characterized purely by their coefficients \(c_k\), meaning there is no need to store discretized values which potentially saves space and computation. The second benefit is that other functions such as polynomials can be approximated by them using the Fourier transform. This means that despite limiting the set of functions to Fourier functions, the behavior of PDEs with other sets of functions can be approximated to a certain extent. The final benefit is the mature ecosystem around these functions which include fast algorithms for the Fourier transform and even numerical approaches for solving known PDEs like the previously mentioned spectral method which we discussed in \lccref{sec:fourier_discussion}. In summary, the generated data consists of features and labels which are Fourier coefficients implicitly defining the relationship enforced by a PDE and scenario. To proceed with the data generation, there are 4 steps involved:
\begin{enumerate}
  \item Scenario and PDE Determination:
        The first step of data generation is determining the scenario related to the PDE or governing equations. First, the PDE to be modeled is determined based on the goals of the dataset. Then, one or more of the parameters is chosen to predict the solution. The chosen parameters will be called the input functions and the solution will be called output functions from here on. The second part of the scenario is the domain or more simply the physical space occupied by the system to be modeled. The domain will be used to compute the function values in relation to the physical space from coefficients of Fourier functions.
  \item Parameter Determination:
        The second step is determining all parameters other than that input parameter based on the scenario and governing equation. These parameters may be coefficients such as material properties like density or viscosity, or forcing terms which model external influence on the system like a heat source in the case of the heat equation. Depending on the parameters, solutions of PDEs may behave very differently, such as the appearance of discontinuities in the solution to low viscosity Burgers' equation. Because of this, the choice of parameters is guided by what the dataset seeks to do.
  \item Random Coefficient Generation:
        The third step of data generation is randomly generating the solutions for the chosen equation. Generating random functions in the space of Fourier functions exploits the fact that the coefficients characterize the function completely. By randomly assigning coefficients \(c_k\), many functions can be generated randomly with very little cost. Since the coefficients need to be complex numbers as in \lccref{eq:complex_number}, both real and imaginary components are generated independently by assigning a random value to each component for each wave number \(k\). They are then put together again into complex numbers.

@@

\subsection{Data Generation: Anti-derivative}
\noindent The first dataset is a simple one dimensional derivative. This was chosen as a simple proof of concept of the ability to solve a differential equation. The equation is related to many real-world problems such as acceleration and speed. One can imagine a train in an ideal world where acceleration is directly translated into speed. When the train accelerates at time \(t \) by some amount \(a \), we can expect the train to have some speed \(u \). In this ideal world, the relationship between speed and acceleration can be modeled with a simple differential \lccref{eq:derivative}. This scenario is found in many real systems albeit often with many more details such as different components of acceleration from friction, drag, gravity, and other factors.
As previously mentioned in \lccref{sec:data_generation}, both velocity \(u \) and acceleration \(a \) are modeled with Fourier series in \lccrefs{eq:fourier_speed,eq:fourier_acceleration} respectively.
\begin{align}
  u\left( t \right)         & = \sum_{k} \hat{u}_k e^{2\pi ikt} \label{eq:fourier_speed}        \\
  a\left( t \right)         & = \sum_{k} \hat{a}_k e^{2\pi ikt} \label{eq:fourier_acceleration}
\end{align}
The domain of the scenario is a two-hour time window. This number was chosen because it is around the ideal length of travel time on high speed rail in comparison to air travel and car travel \autocite{givoniDevelopmentImpactModern2006,wangEfficiencySpatialEquity2019,wrro2236}. This is the first step in generating this dataset.

In the second step, as the derivative \lccref{eq:derivative} does not contain any parameters other than the acceleration which is the input parameter, there are no other parameters to determine. Therefore, the data generation process proceeds to generating coefficients for the speed functions \(\hat{u} \). The coefficients are assigned randomly from a Gaussian distribution with a mean of zero and standard deviation of one. This choice was made such that most wave numbers will have a coefficient of close to zero leaving a sparse set of wave numbers to mostly affect the resulting function. In total, \num{5000} unique functions are generated with \num{100} complex coefficients each.

@@

In the final step, using randomly generated values of \(\hat{a}_k\), the corresponding values of \(\hat{u}_k\) are computed with \lccref{eq:derivative_coeff}. These coefficients are then used to compute the function values inside the domain. The two-hour time window is represented by a grid of \num{500} discrete points. With the coefficients and evaluation points ready, the function values are computed using \lccrefs{eq:fourier_speed,eq:fourier_acceleration}. Next, noise is added to the function values in order to motivate models learning on the dataset to generalize on noise. To also allow evaluation of how well the model performs with different levels of noise, the samples are duplicated into three copies for each high, medium and low noise levels. The function values of each copy is perturbed with noise from a Gaussian distribution with zero mean and standard deviation of some percentage of the average function value standard deviation. The percentages of each high, medium, and low noise levels are 5, 10, and 50 percent. The perturbed function values are then transformed back into their coefficients. An example generated function and its different perturbed versions is shown in \lccref{fig:antiderivative_noise_levels}. The horizontal axis indicates time which is displayed in units of hours. The vertical axis indicates velocity in abstract units.

\begin{figure}[H]
  \centering
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_clean_perturbed_solution.pgf}
    \end{adjustbox}
    \caption{}\label{fig:antiderivative_noise_levels}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_clean.pgf}
    \end{adjustbox}
    \caption{}\label{fig:antiderivative}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/antiderivative_derivative_clean.pgf}
    \end{adjustbox}
    \caption{}\label{fig:derivative}
  \end{subfigure}
  \caption{(\subref{fig:antiderivative_noise_levels}) A generated function with no noise (clean), low noise level (5\%), medium noise level (10\%), and high noise levels (50\%).\ (\subref{fig:antiderivative}) Antiderivative function with no noise.\ (\subref{fig:derivative}) Derivative function with no noise.}
\end{figure}
% TODO: interesting note that the model trains better on function to coefficient task when the antiderivative is computed from the randomly generated derivative than if the derivative is computed from the randomly generated antiderivative. THIS NEEDS A REASON AND ANALYSIS SO MIGHT NOT BE NEEDED IN THIS WRITING. MAYBE A GOOD IDEA FOR A SEPARATE PIECE

\subsection{Data Generation: Burgers' Equation}\label{sec:data_generation_burgers}
\noindent The second dataset generated in this study concerns the Burgers' equation. This equation has been used to model a variety of cases including fluid dynamics, traffic flow, and, shock waves \autocite{bonkileSystematicLiteratureReview2018,orlandiBurgersEquation2000,becBurgersTurbulence2007,jamesonEnergyEstimatesNonlinear2007}. This equation is also used as a base problem for testing the effectivity of numerical methods in solving non-linear PDEs \autocite{barterShockCapturingPDEbased2008,banksNumericalErrorEstimation2012, tabatabaeiImplicitMethodsNumerical2007,bonkileSystematicLiteratureReview2018}. The nonlinear term in the equation creates steep gradients and even shock waves which are discontinuous with low viscosity conditions. These challenges test the stability of numerical solvers. This is the scenario and reason for the choice of this PDE\@. To control the viscosity and therefore the steepness of gradients, the formulation of the Burgers' equation considered in this study is the forced viscous Burgers' equation in one dimension. For a velocity of \(u(x, t)\), viscosity of \(\nu \), and forcing term of \(f(x, t)\), the formulation of the forced Burgers' equation is shown in \lccref{eq:forced_viscous_burgers}.

The domain of we consider here is based on an exact solution of the equation. This is done so that the learned solver model can be tested on the exact solution. There are a number of sources in literature on the exact solution of the Burgers' equation \autocite{bentonTableSolutionsOnedimensional1972,wazwazPartialDifferentialEquations2010,woodExactSolutionBurgers2006}. In this study, we use a specific solution by \textcite{woodExactSolutionBurgers2006} which can be seen in \lccref{eq:burgers_exact_solution}. This solution is periodic in space but not in time. Therefore, the domain in space will be based on the spatial periodicity of the exact solution which is two. As such, the space domain spans from 0 to 2. The time domain, on the other hand, is chosen to span from 0 to 10 such that enough details of the solution is included.
\begin{equation}
  u(x,t) = \frac{2\nu\pi e^{-\pi^2\nu t}\sin(\pi x)}{a+e^{-\pi^2\nu t}\cos(\pi x)} \label{eq:burgers_exact_solution}
\end{equation}
Continuing to the second step, the parameters involved in \lccref{eq:forced_viscous_burgers} are the viscosity \(\nu \) and the forcing term \(f \). Since the forcing term is dependent on the solution in this step, the only parameter to determine is the viscosity. We choose three constant viscosity values which are \num{0.1}, \num{0.01}, and \num{0.0}. These values are chosen so that a variety of behaviors are represented in the dataset from viscous to inviscid flow.

After this, in the third step, the solution functions \(u \) are generated with 8 modes in time and 8 in space and \num{500} unique function samples for each viscosity value. Since we don't factor the viscosity in at this point, for ease of comparison, we ensure the same solutions are generated every time by using a new instance of the generator with the same seed number each time. This results in the solutions being the same for all viscosity and other parameters of the equation are adapted to this.

@@

For the reanalysis dataset, we limit the time range to the years 1995{-}2013 for training and 2020{-}2022 for testing. The specific time ranges for training are chosen to balance the diversity within the data and capacity of the machine the processing will be done on. The testing set time range was chosen to be as close to those shown on the WeatherBench 2 website. The filter for time and variables is shown in \lccref{fig:wb2_retrieval}. The reason for the different time ranges between training and testing is to reduce the risk of information leakage across the training and testing sets. This ensures that the testing scores we measure will be for the most part from the generalization that model learned during training. The filter is an array the length of the number of samples. The array values are all the datetime values that the year ranges are true for. This results in an array of boolean values. Using this boolean filter we can index the data so that we get the filtered data. Then, the filtered data is saved locally so that we can access it in a more performant way. The encoding chunks may need to be removed in order to evade the Xarray library from mistakenly persisting the remote encoding after slicing data \autocite{ianZarrEncodingAttributes}.
\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Python]
  years_idx = ds['time.year']
  data_idx = (years_idx > START_YEAR) & (years_idx < END_YEAR)
  da_orig = ds['2m_temperature'][data_idx]
  # save locally
  da_orig.encoding.pop('chunks',None)
  da_orig.to_zarr(DATA_FILENAME, mode='w')
  # open local version
  da = xr.open_zarr(DATA_FILENAME)
  \end{lstlisting}
  \caption{Example of filtering of Xarray dataset by year and variable. The data is saved and then loaded locally.}\label{fig:wb2_retrieval}
\end{figure}

The dataset includes 62 variables with some variables also spanning 13 discrete vertical levels in the atmosphere. For our use, we will only be using the 2-meter temperature of the atmosphere. This is the air temperature 2 meters above the surface be it land, sea, or inland water. This variable is in units of kelvin (K). Vertically, the values of this variable is obtained by interpolating between the model values at the surface and the lowest model vertical level. An example of the data that was retrieved can be seen in \lccref{fig:wb2_data}. The data shown is in the equiangular projection. This means that towards the poles, the shapes become more distorted. While this is not the most ideal way to represent the data, using the data in this way adds to the challenge of how the model can adapt to the use of suboptimal configurations in the data. To compute the anomaly, we can simply subtract the current state in \lccref{fig:wb2_train_data} by the climatology in \lccref{fig:wb2_clim_data} resulting in \lccref{fig:wb2_train_anomaly}.
\begin{figure}[H]
  \centering
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_train_data.pgf}
    \end{adjustbox}
    \caption{2-meter temperature}\label{fig:wb2_train_data}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_clim_data.pgf}
    \end{adjustbox}
    \caption{Climatology}\label{fig:wb2_clim_data}
  \end{subfigure}
  \begin{subfigure}{\linewidth}
    \begin{adjustbox}{width=\linewidth}
      \input{figures/wb2_train_anomaly.pgf}
    \end{adjustbox}
    \caption{2-meter temperature anomaly}\label{fig:wb2_train_anomaly}
  \end{subfigure}
  \caption{Data retrieved:(\subref{fig:wb2_train_data}) 2-meter temperature at January 1, 1995 6:00 UTC\@. (\subref{fig:wb2_clim_data}) 2-meter temperature climatology for the first day of the year at hour 6 UTC\@. (\subref{fig:wb2_train_anomaly}) Anomaly for 2-meter temperature at January 1, 1995 6:00 UTC\@.}\label{fig:wb2_data}
\end{figure}

After filtering, our training set has the dimensions of \num{27760} time steps, \num{64} longitude, and \num{32} latitude. The testing set on the other hand has \num{4384} time steps and the same values for the other two dimensions. For uses, such as a validation set during hyperparameter optimization, one can use a subset of the training set. The dimensions of the climatology are 4 for hour (0, 6, 12,\& 18), \num{366} for day of year, \num{64} for longitude, and 32 for latitude.

@@

The second use of a dataset depends on whether it can be interpreted as a mapping of a previous function to the next function. In other words, a series of functions related to the previous function as a sum of the previous function and a perturbation. A commonly used form of this are time series. Generally, a time series can be represented as \lccref{eq:time_step_function}. This function describes the evolution of function \(f\) as a sum of some initial state \(f^0\) and perturbation which is its rate of change \(\dv{f}{t}\) at a point in time close to the current time multiplied by the time step \(\delta t\). This is the first order Taylor polynomial. Adding more terms with higher order derivatives can be done to increase the accuracy. This describes the same process that is used in the finite difference method.
\begin{equation}
  f^{t+1} = f^{t} + \delta t \times \dv{f}{t} + \dots \label{eq:time_step_function}
\end{equation}
In this case the previous function and any other independent variables like the forcing term involved in the right-hand side of the equation is used to predict the difference between the future state and the current state. It would mean that each sample function in the data would need to at least have two dimensions. This is because one dimension represents the evolution of the function over time, while the other dimensions define the spatial domain where the function's values are computed at each time step. This condition is true for both the Burgers' equation dataset and the WeatherBench 2 collection of ERA5 datasets. First, the Burgers' equation dataset, while is also able to be used in the first way as a mapping between the forcing term and the solution, one of the traditional ways of solving it is using the method of lines. This entails solving the dimensions other that time with methods such as the finite differences method or the spectral method. The time dimension is then separately solved using FDM such as the form shown in \lccref{eq:time_step_function} \autocite{schiesserNumericalMethodLines2012,sadikuSimpleIntroductionMethod2000}. This is the form that the Burgers' equation dataset will be used in. In this case, algebraic manipulation of \lccref{eq:forced_viscous_burgers} results in the time derivative of the solution as in \lccref{eq:forced_viscous_burgers_time_derivative}.
\begin{align}
  \pdv{u}{t} = f - u\pdv{u}{x} + \nu\pdv[2]{u}{x} \label{eq:forced_viscous_burgers_time_derivative}
\end{align}
The most important component of the right-hand side of the equation, the time derivative, is expressed in terms of the solution function, the forcing term and any other partial derivatives. Because of this, it means we can train the model to learn the relationship between the solution function combined with the forcing term and the time derivative. Putting this together with \lccref{eq:time_step_function}, it is possible to predict the solution function in a future time based on the current solution function and forcing term. The input functions in this case are both the value of the solution function and the forcing term at the current time step. The output function is the time derivative which is computed by rearranging \lccref{eq:time_step_function} resulting in \lccref{eq:time_derivative_time_step_function}.
\begin{equation}
  \dv{f}{t} = \frac{f^{t+1} - f^{t}}{\delta t} \label{eq:time_derivative_time_step_function}
\end{equation}
Perhaps in a more obvious manner, the same is also applied to the WeatherBench 2 collection of ERA5 datasets. The 2-meter temperature is a series of snapshots in time of the distribution of temperature all over the world at 2 meters above the surface. The input function in this case is the temperature distribution at the current time step. The output function is the time derivative of the temperature distribution computed using \lccref{eq:time_derivative_time_step_function}.

The input and output functions are then transformed into basis coefficients using the Fourier transform we discussed in \lccref{sec:fourier_discussion}. Once the coefficients are computed, the input and output sample pairs are split into train, test, and validation subsets. For the first type of input and output function pair, since the samples are independent of each other, a random sampling of the dataset without replacement is done to select the samples for each subset. For the second type of input and output function pairs, if there are multiple samples of functions as in the Burgers' equation dataset, the same process with random sampling is followed. On the other hand, if only one function sample is available, such as the WeatherBench 2 dataset, an issue that arises is the fact that the samples are dependent on one another. This is because a previous time step influences the next time step. This is a problem for generalizing the learned model, since random sampling would lead to the model learning information in the future of the testing subset \autocite{kapoorLeakageReproducibilityCrisis2023,kaufmanLeakageDataMining2012}. As such, the portioning of the dataset is done by dividing the sample into different sections of the function in time.

@@

\section{Implementation of Computational Model}
\noindent This section discusses the implementation of modeling and approximation of PDEs using LSSVR\@. Using the computational model design in \lccref{sec:computational_model_design}, the computational model is implemented as a library using the Python programming language. The development process itself is based on the iterative development model as mentioned in \lccref{sec:development_method}. In this section, the output iterative processes are consolidated and explained together.

@@

\section{Experimental Results}\label{sec:experimental_results}
\noindent In this section, the results of each experimental scenario is presented and analyzed. The results come in the form of testing metrics mentioned in \lccref{sec:evaluation_metrics}. The model interpretations are displays of the correlation image and the p-matrix. The function values and coefficients themselves will also be plotted depending on the number of dimensions.
